Value Ordering, Discrepancies, and Restarts for Subgraph Algorithms

### 2018-07-25 paper/paper.tex
AAAAAAAAAAI

\documentclass[a4paper,UKenglish]{lipics-v2018}

\usepackage{microtype}
\usepackage{complexity}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{tikz}

% \usepackage{showframe}

\bibliographystyle{plainurl}

\newcommand{\neighbourhood}{\operatorname{N}}
\newcommand{\vertexset}{\operatorname{V}}
\newcommand{\nds}{\operatorname{S}}

@@

\newcommand{\figureref}[1]{Figure~\ref{#1}}
\newcommand{\Figureref}[1]{Figure~\ref{#1}}
\newcommand{\twofigureref}[2]{Figures~\ref{#1} and~\ref{#2}}
\newcommand{\algorithmref}[1]{Algorithm~\ref{#1}}
\newcommand{\Algorithmref}[1]{Algorithm~\ref{#1}}

@@

\title{Value Ordering, Discrepancies, and Restarts for Subgraph Algorithms}
\author{Ciaran McCreesh}{University of Glasgow}{ciaran.mccreesh@glasgow.ac.uk}{}{}
\author{Patrick Prosser}{University of Glasgow}{patrick.prosser@glasgow.ac.uk}{}{}
\author{James Trimble}{University of Glasgow}{j.trimble.1@research.gla.ac.uk}{}{}

\authorrunning{C. McCreesh, P. Prosser and J. Trimble}

\Copyright{Ciaran McCreesh, Patrick Prosser and James Trimble}

\subjclass{
\ccsdesc[500]{Mathematics of computing~Graph algorithms},
\ccsdesc[500]{Theory of computation~Discrete optimization},
\ccsdesc[500]{Theory of computation~Backtracking}}

\keywords{Subgraph isomorphism, combinatorial search, rich search}

\funding{This work was supported by the Engineering and Physical Sciences Research Council [grant
numbers EP/026842/1 and EP/M508056/1]}

\acknowledgements{The authors would like to thank Ruth Hoffmann, Christophe Lecoutre, Christine
Solnon and Craig Reilly for their comments.}

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{26th Annual European Symposium on Algorithms (ESA 2018)}
\EventShortTitle{Preprint}
\EventAcronym{ESA}
\EventYear{2018}
\EventDate{August 20--22, 2018}
\EventLocation{Helsinki, Finland}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%\nolinenumbers %uncomment to disable line numbering
\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The subgraph isomorphism problem is to decide whether a copy of a small ``pattern'' graph inside a
larger ``target'' graph. Although \NP-complete, by combining design techniques from artificial
intelligence with careful algorithm engineering, modern subgraph isomorphism solvers can often
produce exact solutions quickly even on graphs with thousands of vertices. The current single
strongest subgraph isomorphism solver uses ``highest degree first'' as a value-ordering heuristic to
direct a constraint programming style search
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:conf/ijcai/McCreeshPT16}. This
heuristic is much better than branching randomly, but is still far from perfect.  We could try to
introduce tiebreaking mechanisms and similar refinements to strengthen the heuristic. However, given
the extreme cost of making an incorrect branching choice early in search, we believe it is more
fruitful to investigate alternatives to simple backtracking (depth-first) search. This paper
evaluates two such approaches: discrepancy search (which performs poorly), and a new form of search
(which is extremely beneficial) based around restarts and nogood recording. To explore different
parts of the search space when restarting, we introduce a novel value-ordering heuristic, which adds
a small amount of randomness to the ``highest degree first'' heuristic.  The effects of these
changes are shown in \figureref{figure:scatter-by-family}: we can make a state-of-the-art algorithm
perform much better on a large number of satisfiable instances, whilst performing worse only rarely
on satisfiable instances, and never on unsatisfiable instances.  Further, these properties are
\emph{not} shared by discrepancy searches (\figureref{figure:scatter-dds}) or by random
value-ordering with restarts (\figureref{figure:scatter-random}). Finally, we show that this
strategy is also effective in an optimisation setting, producing benefits in two maximum common
induced subgraph algorithms.

@@

Let $G$ be a graph. We denote its set of vertices by $\vertexset(G)$, and write $v \sim_G w$ to mean
that vertices $v$ and $w$ are adjacent; we allow a vertex to be adjacent to itself. The
\emph{neighbourhood} of a vertex, $\neighbourhood_G(v)$, is the set of vertices adjacent to $v$, and
the \emph{degree} of $v$, $\deg_G(v)$, is the cardinality of its neighbourhood. The
\emph{neighbourhood degree sequence} of $v$, $\nds_G(v)$, is the sequence of degrees of the
neighbours of $v$ in descending order, and we denote by $\preceq$ the inclusion relation of
Zampelli et al.\ \cite{DBLP:journals/constraints/ZampelliDS10}. When the graph is clear from the context, we omit
the $G$ subscripts. As per Hoffmann et al.\ \cite{DBLP:conf/aaai/HoffmannMR17}, $G^{n,\ell}$ denotes the graph with
vertex set $\vertexset(G)$, and with edges between vertices $v$ and $w$ if there are at least $n$
simple paths of length exactly $\ell$ between vertices $v$ and $w$ in $G$.

@@

We now briefly outline the non-shaded parts of \algorithmref{algorithm:sip}, which is how we will solve the
subgraph isomorphism problem. This algorithm is very closely based upon the $k{\downarrow}$
algorithm of Hoffmann et al.\ \cite{DBLP:conf/aaai/HoffmannMR17} with $k = 0$, and we refer the reader to that paper
for full technical details; that algorithm, in turn, is a simplification of the Glasgow2 algorithm
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}.

\Algorithmref{algorithm:sip} works by recursively building up a valid mapping of pattern vertices to target
vertices. For each pattern vertex $v$, we have a domain $D_v$ containing a set of feasible target
vertices. Each domain is initialised at the top of search (\siplinerangeref{initstart}{initend}),
using the invariants described by Zampelli et al.\ \cite{DBLP:journals/constraints/ZampelliDS10} and
Hoffmann et al.\ \cite{DBLP:conf/aaai/HoffmannMR17} to restrict its initial values.

@@

Our $\FuncSty{propagate}$ routine is also close to that of Hoffmann et al.\ \cite{DBLP:conf/aaai/HoffmannMR17},
differing only in that we carry out the filtering for injectivity
(\siplinerangeref{alldiffstart}{alldiffend}) after every assignment rather than once at the end of
the unit propagation loop. (We make this change because the all-different propagator used is not
idempotent, and we wish to avoid an odd interaction when we introduce nogood filtering.)

@@

        (\mathcal{P}^{2,2},\ \mathcal{T}^{2,2}), (\mathcal{P}^{3,2},\ \mathcal{T}^{3,2}),

@@

        \nl \lForEach{$(P,\,T) \in L$}{$D_v \gets \{ w \in D_v : $ \siplabel{degreefiltering} $v \sim_P v \Rightarrow w \sim_T w~\wedge$
            $\nds_P(v) \preceq \nds_T(w)\}$\siplabel{initend}

@@

            \nl $S \gets \FuncSty{search}(\{ E \in D'\,{:}\,\left|E\right|{>}\,1 \}, B \wedge (v\,{\mapsto}\,v'))$\siplabel{recurse} \;

@@

            \nl \lForEach{$(P,\,T) \in L~\KwSty{where}~v \sim_P w$}{$D_w \gets D_w \cap \neighbourhood_T(v')$}

@@

\nl \ForEach{$D_v \in D$ \textnormal{from smallest cardinality upwards\siplabel{eachdomain}}}{

@@

For the variable-ordering heuristic on \siplineref{variableordering} of
\algorithmref{algorithm:sip}, we use smallest domain first, tiebreaking on highest degree
\cite{DBLP:journals/ai/HaralickE80,DBLP:journals/jair/McCreeshPST18}. Variable ordering is not the
focus of this paper---as far as possible, we use this same dynamic ordering throughout the paper, so
as not to affect the performance of the algorithm on unsatisfiable instances.  For value ordering,
which occurs on \siplineref{valueordering} of \algorithmref{algorithm:sip}, we will initially select
vertices from highest degree to lowest degree; we return to value ordering in the second half of
this paper. Note that the variable-ordering heuristic affects both unsatisfiable and satisfiable
instances, whilst the value-ordering heuristic is relevant only for satisfiable instances.

@@

Our experiments are performed on systems with dual Intel Xeon E5-2697A v4 CPUs and 512GBytes RAM,
running Ubuntu 17.04. We
implemented\footnote{\url{https://github.com/ciaranm/restarting-sip-paper}, DOI to be issued pre-publication}
\algorithmref{algorithm:sip} in C++ using the bit-parallel implementation of
Hoffmann et al.\ \cite{DBLP:conf/aaai/HoffmannMR17} as a starting point, and compiled it using GCC 6.3.0. We use a
deterministic pseudo-random number generator for reproducibility.

We use the dataset introduced by Kotthoff et al.\ \cite{DBLP:conf/lion/KotthoffMS16} for evaluation. This dataset
brings together a range of randomly-generated and application instance families from earlier papers:

@@

We begin by verifying that our implementation of the basic \algorithmref{algorithm:sip} is competitive. In
\figureref{figure:others} we show the cumulative number of instances solved over time, using our
implementation (both with the degree heuristic, and with the modifications described in the
remainder of this paper), the Glasgow2 and Glasgow3 algorithms from which \algorithmref{algorithm:sip} is derived
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}, the PathLAD variation of the LAD
algorithm \cite{DBLP:journals/ai/Solnon10,DBLP:conf/lion/KotthoffMS16}, and VF2
\cite{DBLP:journals/pami/CordellaFSV04}, in each case using the original implementation provided
by the algorithm's authors. For each algorithm, the $y$ value
gives the cumulative number of instances which (individually) can be solved in no more than $x$
milliseconds.  The vertical distance between two lines therefore shows how many more instances can
be solved by one solver than another, if every instance is run separately with the chosen $x$
timeout. The horizontal distance shows how many times longer the per-instance timeout would need to
be to allow the rightmost algorithm to succeed on $y$ out of the 14,621 instances (bearing in mind
that the two sets of $y$ instances could be different).

As expected, the performance of \algorithmref{algorithm:sip} with the degree heuristic is very close to that
of Glasgow2, although we perform better at lower runtimes because we use simpler preprocessing at
the top of search. Our implementation also clearly beats PathLAD and VF2, except for very low
choices of timeout.

The dataset includes many instances which are extremely easy for a good solver, and so it can be
hard to see the differences between the stronger solvers at higher runtimes. This paper focusses
upon improving the performance on the remaining hard satisfiable instances, and so in the second
plot in \figureref{figure:others} (and in subsequent cumulative plots) we show only satisfiable
instances, and use a reduced range on both axes.

@@

\Figureref{figure:value-ordering-heuristics} shows the cumulative number of satisfiable instances
solved over time using different value-ordering heuristics:

@@

The performance of the degree and biased heuristics are very similar: the
biased heuristic has slightly worse performance below a 400s timeout, and effectively equal
performance above it. This shows that we can introduce an element of randomness into the degree
value-ordering heuristic without adversely affecting its performance \emph{in aggregate}.
\Figureref{figure:scatter-heuristics} gives a more detailed look at these results. This scatter plot
shows that despite the aggregate performance being very similar, on a case by case basis, the two
heuristics can make a large difference to the performance of satisfiable instances.

@@

\Figureref{figure:scatter-dds} shows the effects of adding Walsh's \cite{DBLP:conf/ijcai/Walsh97}
depth-bounded discrepancy search (DDS) to \algorithmref{algorithm:sip}. Each point represents the solving
time for one instance---to avoid noise for easier instances, we measure only time spent during
search, and exclude time spent in preprocessing and initialisation.  Points below the $x-y$ diagonal
are speedups, whilst points on the top and right axes represent instances which timed out after one
thousand seconds with one algorithm, but not the other. For satisfiable instances, the different
point styles show the different families, whilst all unsatisfiable instances are shown as dark dots.

@@

\subsection{Adding Restarts and Nogood Recording to the Algorithm}

Another alternative to plain backtracking search is provided by \emph{restarts}. The general idea is
to perform a certain amount of search, and then if no solution has been found (and unsatisfiability
has not been proven), to abandon search and restart from the beginning. Obviously, such an approach
can only be beneficial if something changes after restarting---in a constraint programming setting,
this is usually the variable-ordering heuristic
\cite{DBLP:journals/jsat/LecoutreSTV07,DBLP:conf/cp/GayHLS15,DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17}.
In this paper, we instead rely upon randomness in the \emph{value}-ordering heuristic, and continue
to use smallest domain first with static tiebreaking for variable-ordering.\footnote{It may be
possible to further improve \algorithmref{algorithm:sip} by also introducing randomness or some form of
learning into its variable-ordering heuristic. However, simultaneously introducing a second change
would considerably complicate the empirical analysis.} Using restarts on value-ordering heuristics
appears to be uncommon---although Razgon et al.\ \cite{DBLP:conf/flairs/RazgonOP07} look at learning
value-ordering heuristics from restarts, Chu et al.\ \cite{DBLP:conf/cp/ChuSS09} use a similar
scheme in the context of parallel search, and an early approach by Gomes et
al.\ \cite{DBLP:conf/aaai/GomesSK98} does so in an optimisation context.

The shaded code shows how to introduce restarts to \algorithmref{algorithm:sip}. The recursive
$\FuncSty{search}$ procedure is modified to return either true (if the instance is satisfiable),
false (unsatisfiable), or a third value if some threshold was reached without the solution being
known.  The main function then calls $\FuncSty{search}$ in a loop (\siplineref{restartsloop}), until
a solution is found.

@@

\begin{figure}[p]
    \centering
    \includegraphics*{gen-graph-restarts.pdf}

@@

\begin{figure}[p]
    \includegraphics*{gen-graph-scatter-random-goods.pdf}
    \hfill
    \includegraphics*{gen-graph-scatter-random.pdf}
    \caption{Comparing basic backtracking with the degree heuristic, versus the random heuristic
    with restarts. The plot on the right uses nogood recording, whilst the plot on the left does not.}
    \label{figure:scatter-random}
\end{figure}

\begin{figure}[p]
    \centering
    \includegraphics*{gen-graph-scatter-by-family.pdf}

    \caption{Comparing basic backtracking with the degree heuristic, versus the biased
    heuristic with restarts and nogood recording.}
    \label{figure:scatter-by-family}
\end{figure}

In \figureref{figure:restarts} we show the effects of adding restarts and nogood recording to the
algorithm. With restarts and nogood recording,
the random value-ordering heuristic comfortably beats the degree strategy with simple backtracking
search. In other
words, although having a good value-ordering heuristic is beneficial, introducing randomness into
the search is better, if it is done alongside a mechanism to avoid heavy commitment to any
particular random choice. \Figureref{figure:scatter-random} looks at
this in more detail. On the left-hand plot, without nogood recording, unsatisfiable instances
experience a slowdown of more than an order of magnitude, whilst on the right-hand plot the
performance remains unchanged (which shows that the overheads of nogoods and restarting are
negligible). We therefore use nogood recording in all subsequent experiments with restarts. However,
although better in aggregate on satisfiable instances, on a case by case basis many instances
perform substantially worse with randomisation (either with or without restarts).

\Figureref{figure:restarts} also shows that the biased heuristic together with restarts is better
still---that is, if we are introducing restarts, then it is better to add a small amount of
randomness to a tailored deterministic heuristic than it is to throw away the heuristic altogether.
Indeed, the original algorithm can solve 1,983 satisfiable instances by 993.0 seconds, whilst the
biased and random restarting algorithms require only 9.2 seconds and 33.8 seconds respectively to
solve the same number.

In the more detailed view in \figureref{figure:scatter-by-family}, comparing the basic algorithm with the
degree heuristic to the degree-biased algorithm with restarts, all of the unsatisfiable instances
are very close to the $x-y$ diagonal, showing that their performance is nearly unchanged. On the
other hand, there are large numbers of satisfiable instances well below the diagonal line,
indicating large speedups.  Better yet, there are only a handful of satisfiable instances that are
more than a factor of ten times worse. In other words, as well as improving performance, we have
made up most of the consistency we lost by introducing randomness.

@@

\begin{figure}[tb]
    \centering
    \includegraphics*{gen-graph-mcs.pdf}
    \caption{The cumulative number of maximum common induced subgraph instances solved by
    k${\downarrow}$ and McSplit${\downarrow}$ over time, with and without restarts and nogood
    recording.}\label{figure:mcs}
\end{figure}

\begin{figure}[tb]
    \includegraphics*{gen-graph-scatter-kdown.pdf}
    \hfill
    \includegraphics*{gen-graph-scatter-mcsplit.pdf}
    \caption{On the left, comparing k${\downarrow}$ with and without restarts and nogood recording on an instance
    by instance basis. On the right, the same, for McSplit${\downarrow}$.}\label{figure:mcsplitandkdown}
\end{figure}

Having looked at subgraph isomorphism in detail, we now briefly discuss the more general maximum
common induced subgraph problem.  Two recent algorithms for this problem also make use of
backtracking search with degree as a value-ordering heuristic. The k${\downarrow}$ algorithm
\cite{DBLP:conf/aaai/HoffmannMR17} attempts to solve the problem by first trying to solve the
induced subgraph isomorphism problem, and then if that fails, retries allowing a single unmatched
vertex (and thus using weaker invariants), and so on. Due to its similarity to \algorithmref{algorithm:sip},
we can introduce the same bias and restart strategy, requiring only small changes
to handle wildcards in nogoods.

@@

Performance results from these two modified algorithms, using the same families of instances as in
the previous section, are shown in \twofigureref{figure:mcs}{figure:mcsplitandkdown}. Although we have
moved from a decision problem to an optimisation problem, the same changes remain clearly
beneficial. For the $k{\downarrow}$ algorithm, the change has a minimal effect on many instances
(typically, where the $k = 0$ subproblem is unsatisfiable and hard, and the $k = 1$ subproblem is
satisfiable and easy), but gives large benefits on many more instances than it penalises: it is over
an order of magnitude better on nearly a hundred instances, whilst being an order of magnitude worse
on only seven.

@@

\bibliography{dblp}

>>>>

\def\year{2019}\relax
%File: formatting-instruction.tex
\documentclass[letterpaper]{article} %DO NOT CHANGE THIS
\usepackage{aaai18}  %Required
\usepackage{times}  %Required
\usepackage{helvet}  %Required
\usepackage{courier}  %Required
\usepackage{url}  %Required
\usepackage{graphicx}  %Required
\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:
  \pdfinfo{
/Title (Value Ordering, Discrepancies, and Restarts for Subgraph Algorithms)
/Author (Anonymous and Anonymous and Anonymous)}
\setcounter{secnumdepth}{0}

\usepackage{complexity}
\usepackage{amsmath}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{tikz}
\usepackage{cleveref}

\usepackage{showframe}

\newcommand{\neighbourhood}{\operatorname{N}}
\newcommand{\vertexset}{\operatorname{V}}

@@

\newcommand{\citet}[1]{\citeauthor{#1} \shortcite{#1}}

@@

\title{Value Ordering, Discrepancies, and Restarts for Subgraph Algorithms}
\author{Anonymous \and Anonymous \and Anonymous \\ email address \\ address}

@@

The subgraph isomorphism problem is to decide whether a copy of a small ``pattern'' graph inside a
larger ``target'' graph. Although \NP-complete, by combining design techniques from artificial
intelligence with careful algorithm engineering, modern subgraph isomorphism solvers can often
produce exact solutions quickly even on graphs with thousands of vertices. The current single
strongest subgraph isomorphism solver uses ``highest degree first'' as a value-ordering heuristic to
direct a constraint programming style search
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:conf/ijcai/McCreeshPT16}. This
heuristic is much better than branching randomly, but is still far from perfect.  We could try to
introduce tiebreaking mechanisms and similar refinements to strengthen the heuristic. However, given
the extreme cost of making an incorrect branching choice early in search, we believe it is more
fruitful to investigate alternatives to simple backtracking (depth-first) search. This paper
evaluates two such approaches: discrepancy search (which performs poorly), and a new form of search
(which is extremely beneficial) based around restarts and nogood recording. To explore different
parts of the search space when restarting, we introduce a novel value-ordering heuristic, which adds
a small amount of randomness to the ``highest degree first'' heuristic.  The effects of these
changes are shown in the final plot of \cref{figure:scatter-random}: we can make a state-of-the-art algorithm
perform much better on a large number of satisfiable instances, whilst performing worse only rarely
on satisfiable instances, and never on unsatisfiable instances.  Further, these properties are
\emph{not} shared by discrepancy searches (\cref{figure:scatter-dds}) or by random
value-ordering with restarts (the first two plots of \cref{figure:scatter-random}). Finally, we
show that this strategy is also effective in an optimisation setting, producing benefits in two
maximum common induced subgraph algorithms.

@@

Let $G$ be a graph. We denote its set of vertices by $\vertexset(G)$, and write $v \sim_G w$ to mean
that vertices $v$ and $w$ are adjacent; we allow a vertex to be adjacent to itself. The
\emph{neighbourhood} of a vertex, $\neighbourhood_G(v)$, is the set of vertices adjacent to $v$, and
the \emph{degree} of $v$, $\deg_G(v)$, is the cardinality of its neighbourhood. When the graph is clear from the context, we omit
the $G$ subscripts. As per \citet{DBLP:conf/aaai/HoffmannMR17}, $G^{n,\ell}$ denotes the graph with
vertex set $\vertexset(G)$, and with edges between vertices $v$ and $w$ if there are at least $n$
simple paths of length exactly $\ell$ between vertices $v$ and $w$ in $G$. Finally, we define a
function $\FuncSty{compatible}(v, w)$ to mean that mapping pattern vertex $v$ to target vertex $w$
is feasible based upon both loop constraints and the neighbourhood degree sequence relation of
\citet{DBLP:journals/constraints/ZampelliDS10}.

@@

We now briefly outline the non-shaded parts of \cref{algorithm:sip}, which is how we will solve the
subgraph isomorphism problem. This algorithm is very closely based upon the $k{\downarrow}$
algorithm of \citet{DBLP:conf/aaai/HoffmannMR17} with $k = 0$, and we refer the reader to that paper
for full technical details; that algorithm, in turn, is a simplification of the Glasgow2 algorithm
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}.

\Cref{algorithm:sip} works by recursively building up a valid mapping of pattern vertices to target
vertices. For each pattern vertex $v$, we have a domain $D_v$ containing a set of feasible target
vertices. Each domain is initialised at the top of search (\siplinerangeref{initstart}{initend}),
using the invariants described by \citet{DBLP:journals/constraints/ZampelliDS10} and
\citet{DBLP:conf/aaai/HoffmannMR17} to restrict its initial values.

@@

Our $\FuncSty{propagate}$ routine is also close to that of \citet{DBLP:conf/aaai/HoffmannMR17},
differing only in that we carry out the filtering for injectivity
(\siplinerangeref{alldiffstart}{alldiffend}) after every assignment rather than once at the end of
the unit propagation loop. (We make this change because the all-different propagator used is not
idempotent, and we wish to avoid an odd interaction when we introduce nogood filtering.)

@@

        (\mathcal{P}^{2,2},\ \mathcal{T}^{2,2}),$ \\
        $\hspace*{3em}(\mathcal{P}^{3,2},\ \mathcal{T}^{3,2}),

@@

        \nl \ForEach{$(P,\,T) \in L$}{\nl$D_v \gets \{ w \in D_v : \FuncSty{compatible}(v, w)\}$\siplabel{initend}

@@

            \nl $S \gets \FuncSty{search}(\{ E \in D'\,{:}\,\left|E\right|{>}\,1 \}, B \wedge (v\,{\mapsto}\,v'))$\siplabel{recurse}\;

@@

            \nl \ForEach{$(P,\,T) \in L~\KwSty{where}~v \sim_P w$}{\nl$D_w \gets D_w \cap \neighbourhood_T(v')$}

@@

\nl \ForEach{$D_v \in D$ \textnormal{in ascending cardinality order\siplabel{eachdomain}}}{

@@

For the variable-ordering heuristic on \siplineref{variableordering} of
\cref{algorithm:sip}, we use smallest domain first, tiebreaking on highest degree
\cite{DBLP:journals/ai/HaralickE80,DBLP:journals/jair/McCreeshPST18}. Variable ordering is not the
focus of this paper---as far as possible, we use this same dynamic ordering throughout the paper, so
as not to affect the performance of the algorithm on unsatisfiable instances.  For value ordering,
which occurs on \siplineref{valueordering} of \cref{algorithm:sip}, we will initially select
vertices from highest degree to lowest degree; we return to value ordering in the second half of
this paper. Note that the variable-ordering heuristic affects both unsatisfiable and satisfiable
instances, whilst the value-ordering heuristic is relevant only for satisfiable instances.

@@

Our experiments are performed on systems with dual Intel Xeon E5-2697A v4 CPUs and 512GBytes RAM,
running Ubuntu 17.04. We
implemented\footnote{DOI removed for anonymous review}
\cref{algorithm:sip} in C++ using the bit-parallel implementation of
\citet{DBLP:conf/aaai/HoffmannMR17} as a starting point, and compiled it using GCC 6.3.0. We use a
deterministic pseudo-random number generator for reproducibility.

We use the dataset introduced by \citet{DBLP:conf/lion/KotthoffMS16} for evaluation. This dataset
brings together a range of randomly-generated and application instance families from earlier papers:

@@

We begin by verifying that our implementation of the basic \cref{algorithm:sip} is competitive. In
\cref{figure:others} we show the cumulative number of instances solved over time, using our
implementation (both with the degree heuristic, and with the modifications described in the
remainder of this paper), the Glasgow2 and Glasgow3 algorithms from which \cref{algorithm:sip} is derived
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}, the PathLAD variation of the LAD
algorithm \cite{DBLP:journals/ai/Solnon10,DBLP:conf/lion/KotthoffMS16}, and VF2
\cite{DBLP:journals/pami/CordellaFSV04}, in each case using the original implementation provided
by the algorithm's authors. For each algorithm, the $y$ value
gives the cumulative number of instances which (individually) can be solved in no more than $x$
milliseconds.  The vertical distance between two lines therefore shows how many more instances can
be solved by one solver than another, if every instance is run separately with the chosen $x$
timeout. The horizontal distance shows how many times longer the per-instance timeout would need to
be to allow the rightmost algorithm to succeed on $y$ out of the 14,621 instances (bearing in mind
that the two sets of $y$ instances could be different).

As expected, the performance of \cref{algorithm:sip} with the degree heuristic is very close to that
of Glasgow2, although we perform better at lower runtimes because we use simpler preprocessing at
the top of search. Our implementation also clearly beats PathLAD and VF2, except for very low
choices of timeout.

The dataset includes many instances which are extremely easy for a good solver, and so it can be
hard to see the differences between the stronger solvers at higher runtimes. This paper focusses
upon improving the performance on the remaining hard satisfiable instances, and so in the second
plot in \cref{figure:others} (and in subsequent cumulative plots) we show only satisfiable
instances, and use a reduced range on both axes.

@@

\Cref{figure:value-ordering-heuristics} shows the cumulative number of satisfiable instances
solved over time using different value-ordering heuristics:

@@

The performance of the degree and biased heuristics are very similar: the
biased heuristic has slightly worse performance below a 400s timeout, and effectively equal
performance above it. This shows that we can introduce an element of randomness into the degree
value-ordering heuristic without adversely affecting its performance \emph{in aggregate}.
\Cref{figure:scatter-heuristics} gives a more detailed look at these results. This scatter plot
shows that despite the aggregate performance being very similar, on a case by case basis, the two
heuristics can make a large difference to the performance of satisfiable instances.

@@

\Cref{figure:scatter-dds} shows the effects of adding Walsh's \cite{DBLP:conf/ijcai/Walsh97}
depth-bounded discrepancy search (DDS) to \cref{algorithm:sip}. Each point represents the solving
time for one instance---to avoid noise for easier instances, we measure only time spent during
search, and exclude time spent in preprocessing and initialisation.  Points below the $x-y$ diagonal
are speedups, whilst points on the top and right axes represent instances which timed out after one
thousand seconds with one algorithm, but not the other. For satisfiable instances, the different
point styles show the different families, whilst all unsatisfiable instances are shown as dark dots.

@@

\subsection{Adding Restarts and Nogood Recording}

Another alternative to plain backtracking search is provided by \emph{restarts}. The general idea is
to perform a certain amount of search, and then if no solution has been found (and unsatisfiability
has not been proven), to abandon search and restart from the beginning. Obviously, such an approach
can only be beneficial if something changes after restarting---in a constraint programming setting,
this is usually the variable-ordering heuristic
\cite{DBLP:journals/jsat/LecoutreSTV07,DBLP:conf/cp/GayHLS15,DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17}.
In this paper, we instead rely upon randomness in the \emph{value}-ordering heuristic, and continue
to use smallest domain first with static tiebreaking for variable-ordering.\footnote{It may be
possible to further improve \cref{algorithm:sip} by also introducing randomness or some form of
learning into its variable-ordering heuristic. However, simultaneously introducing a second change
would considerably complicate the empirical analysis.} Using restarts on value-ordering heuristics
appears to be uncommon---although \citet{DBLP:conf/flairs/RazgonOP07} look at learning
value-ordering heuristics from restarts, \citet{DBLP:conf/cp/ChuSS09} use a similar
scheme in the context of parallel search, and an early approach by \citet{DBLP:conf/aaai/GomesSK98} does so in an optimisation context.

The shaded code shows how to introduce restarts to \cref{algorithm:sip}. The recursive
$\FuncSty{search}$ procedure is modified to return either true (if the instance is satisfiable),
false (unsatisfiable), or a third value if some threshold was reached without the solution being
known.  The main function then calls $\FuncSty{search}$ in a loop (\siplineref{restartsloop}), until
a solution is found.

@@

\begin{figure*}[tb]
    \includegraphics*{gen-graph-scatter-random-goods.pdf}
    \hfill
    \includegraphics*{gen-graph-scatter-random.pdf}
    \hfill
    \includegraphics*{gen-graph-scatter-by-family.pdf}
    \caption{Comparing basic backtracking with the degree heuristic, versus the random and biased
    heuristics with restarts. The first plot does not use nogood recording, whilst the second and
    third do.}
    \label{figure:scatter-random}
\end{figure*}

\begin{figure}[tb]
    \centering
    \includegraphics*{gen-graph-restarts.pdf}

@@

In \cref{figure:restarts} we show the effects of adding restarts and nogood recording to the
algorithm. With restarts and nogood recording,
the random value-ordering heuristic comfortably beats the degree strategy with simple backtracking
search. In other
words, although having a good value-ordering heuristic is beneficial, introducing randomness into
the search is better, if it is done alongside a mechanism to avoid heavy commitment to any
particular random choice. \Cref{figure:scatter-random} looks at
this in more detail. On the first plot, without nogood recording, unsatisfiable instances
experience a slowdown of more than an order of magnitude, whilst on the second plot the
performance remains unchanged (which shows that the overheads of nogoods and restarting are
negligible). We therefore use nogood recording in all subsequent experiments with restarts. However,
although better in aggregate on satisfiable instances, on a case by case basis many instances
perform substantially worse with randomisation (either with or without restarts).

\Cref{figure:restarts} also shows that the biased heuristic together with restarts is better
still---that is, if we are introducing restarts, then it is better to add a small amount of
randomness to a tailored deterministic heuristic than it is to throw away the heuristic altogether.
Indeed, the original algorithm can solve 1,983 satisfiable instances by 993.0 seconds, whilst the
biased and random restarting algorithms require only 9.2 seconds and 33.8 seconds respectively to
solve the same number.

In the more detailed view in the final plot of \cref{figure:scatter-random}, comparing the
basic algorithm with the degree heuristic to the degree-biased algorithm with restarts, all of the
unsatisfiable instances are very close to the $x-y$ diagonal, showing that their performance is
nearly unchanged. On the other hand, there are large numbers of satisfiable instances well below the
diagonal line, indicating large speedups.  Better yet, there are only a handful of satisfiable
instances that are more than a factor of ten times worse. In other words, as well as improving
performance, we have made up most of the consistency we lost by introducing randomness.

@@

Having looked at subgraph isomorphism in detail, we now briefly discuss the more general maximum
common induced subgraph problem.  Two recent algorithms for this problem also make use of
backtracking search with degree as a value-ordering heuristic. The k${\downarrow}$ algorithm
\cite{DBLP:conf/aaai/HoffmannMR17} attempts to solve the problem by first trying to solve the
induced subgraph isomorphism problem, and then if that fails, retries allowing a single unmatched
vertex (and thus using weaker invariants), and so on. Due to its similarity to \cref{algorithm:sip},
we can introduce the same bias and restart strategy, requiring only small changes
to handle wildcards in nogoods.

@@

Performance results from these two modified algorithms, using the same families of instances as in
the previous section, are shown in \cref{figure:mcs,figure:kdown,figure:mcsplit}. Although we have
moved from a decision problem to an optimisation problem, the same changes remain clearly
beneficial. For the $k{\downarrow}$ algorithm, the change has a minimal effect on many instances
(typically, where the $k = 0$ subproblem is unsatisfiable and hard, and the $k = 1$ subproblem is
satisfiable and easy), but gives large benefits on many more instances than it penalises: it is over
an order of magnitude better on nearly a hundred instances, whilst being an order of magnitude worse
on only seven.

\begin{figure}[p]
    \centering
    \includegraphics*{gen-graph-mcs.pdf}
    \caption{The cumulative number of maximum common induced subgraph instances solved by
    k${\downarrow}$ and McSplit${\downarrow}$ over time, with and without restarts and nogood
    recording.}\label{figure:mcs}
\end{figure}

\begin{figure}[p]
    \centering
    \includegraphics*{gen-graph-scatter-kdown.pdf}
    \caption{Comparing k${\downarrow}$ with and without restarts and nogood recording on an instance
    by instance basis.}\label{figure:kdown}
\end{figure}

\begin{figure}[p]
    \centering
    \includegraphics*{gen-graph-scatter-mcsplit.pdf}
    \caption{Comparing McSplit${\downarrow}$ with and without restarts and nogood recording on an instance
    by instance basis.}\label{figure:mcsplit}
\end{figure}

@@

\section*{Acknowledgements}

% The authors would like to thank Ruth Hoffmann, Christophe Lecoutre, Christine Solnon and Craig
% Reilly for their comments.  This work was supported by the Engineering and Physical Sciences
% Research Council [grant numbers EP/026842/1 and EP/M508056/1].

The authors would like to thank Anonymous, Anonymous, Anonymous and Anonymous for their comments.
This work was supported by an anonymous funding agency.

\bibliographystyle{aaai}
\bibliography{dblp}
